\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-09-29 chapters/conclusion.tex
Started patients paper

\todo{https://sci-hub.ru/https://ieeexplore.ieee.org/abstract/document/1642754}
\todo{https://www.sciencedirect.com/science/article/abs/pii/B9780444639349000020}
\todo{BCI principles and practice chapter 19}

@@

@@

\section{Recommendations}
\subsection{Working with patients}
start from patients
use all available signals
\subsection{Linear decoding methods}

>>>>

\todo{https://sci-hub.ru/https://ieeexplore.ieee.org/abstract/document/1642754}
\todo{https://www.sciencedirect.com/science/article/abs/pii/B9780444639349000020}
\todo{BCI principles and practice chapter 19}
\todo{https://www.nature.com/articles/s44222-024-00239-5}

@@

\todo{. Coincidentally, this also
builds towards a solution for the Midas Touch Problem in BCI, where a BCI user sometimes accidentally
selects a target while not wanting to give any input. Decoding of mental attention independent of eye
gaze, with the option of having visual attention without mental attention,
would counteract this.}

@@

\section{Recommendations}
\subsection{Working with patients}
immediately move to patients and on-line implementation, to be able to report
interpretable metrics
permanent working collaboration as a lab

start from patients
use all available signals
\subsection{Linear decoding methods}

### 2024-09-29 chapters/gaze_independence.tex
Started patients paper

@@

\section{The gaze-dependence problem}
\label{sec:gaze-dependence}

@@

\begin{table}[h]
  \sffamily
  \footnotesize
	\centering
	\begin{tabular}{@{}l|rrrrr|r@{}}
    & \bfseries \acs{als} & \bfseries \acs{ms}   & \bfseries Stroke &\bfseries\acs{dmd} &\bfseries \acs{sma} &\bfseries \acs{lis} \\ \hline
		\bfseries Minor    & 50\% & 31\% & 40-70\% & + & - &      \\
		\bfseries Severe   & 33\% & 3\%  & +       & - & - & 98\% \\
		\bfseries Complete & 17\% & -    & +       & - & - & 2\%  \\
	\end{tabular}
	\caption{Incidence of eye motor impairment in selected \ac{bci} target
    patient populations. (\acs{als}: \acl{als}, \acs{ms}: \acl{ms}, \acs{dmd}:
    \acl{dmd}, \acs{sma}: \acl{sma}, \acs{lis}: \acl{lis}).
    $+$: frequent, $-$: infrequent.}
    \label{tab:incidence}
\end{table}
\todo{include FA}

@@

In this work, we will use the term \emph{gaze-independent} meaning
\emph{‘dealing explicitly with the fact that a user cannot control their
gaze.'}
In the context of a visual \ac{bci}, this means that the user's
visuospatial attention and their gaze do not necessarily coincide.
For patients with eye motor impairment, gazing directly at a screen target may
be uncomfortable, impractical, or even impossible.
Hence, assistive devices that rely on eye tracking are often inefficient for
them.
Consequently, while \ac{bci}s hold great promise for these individuals, conventional
gaze-dependent \ac{bci} solutions do not meet their needs due to the absence of gaze
control. Therefore, the development of decoding strategies that account for covert
VSA becomes crucial in the pursuit of high-performance gaze independent
\ac{bci}s.

@@

Gaze-independent \ac{erp}-based \ac{bci}s~\cite{Riccio2012, Aloise2012} can be realized in three
ways.
Firstly, active \ac{bci} communication paradigms relying on endogenous activation from the user
do not rely on sensory stimulation.
Examples of this are imagined movement or imagined speech paradigms.
These paradigms can yield very high information transfer
rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
complexity that can be captured in commands, but often only do so when paired
with invasive recording.
Non-visual reactive paradigms that use e.g. auditory and somatosensory
stimulation do not rely on gaze redirection exist but also result in lower information transfer
rates compared to visual paradigms.
Furthermore, they can suffer from increased mental effort in operation and user-dependent variability~\cite{Reichert2020b}.
\todo{insert table from riccio in text as numbers, not as table}

@@

First, we designed a visual oddball interface that helps us study and adapt
to the effects of eye motor impairment on the \ac{erp}, when using existing state-of-the-art
decoders and on our own proposed decoders.
Using a hexagonal lay-out interface, similar to the visual Hex-o-Spell proposed
\textcite{Treder2010}, we present six flashing circular targets
(without letters or symbols) to the participant while the EEG and
electro-oculogram (EOG) are recorded, as well as eye gaze using eye tracking.

>>>>

\epigraph{%
  ``\elide\ This begs the question: if a patient had reliable control
  over their eye gaze, why would one aim to facilitate communication with that
  patient via a BCI? It would be far simpler and more efficient to use an
  eye-tracking system.
  And even if a BCI were the preferred or only available option for some reason,
  designs requiring eye gaze control could not be used
  by patients with severe motor impairment, as these patients
  cannot direct their gaze.
  Because these are the patients who have the greatest need for BCIs, it is
  important that accurate gaze-independent BCIs exist.''
}{%
  \textcite{Egan2017}
}

@@

\section{The gaze-dependence problem}%
\label{sec:gaze-dependence}

@@

\begin{table}[h]
  \sffamily
  \footnotesize
	\centering
	\begin{tabular}{@{}l|rrrrr|r@{}}
    & \bfseries \acs{als} & \bfseries \acs{ms}   & \bfseries Stroke &\bfseries\acs{dmd} &\bfseries \acs{sma} &\bfseries \acs{lis} \\ \hline
		\bfseries Minor    & 50\% & 31\% & 40-70\% & + & - &      \\
		\bfseries Severe   & 33\% & 3\%  & +       & - & - & 98\% \\
		\bfseries Complete & 17\% & -    & +       & - & - & 2\%  \\
	\end{tabular}
	\caption{Incidence of eye motor impairment in selected \ac{bci} target
    patient populations. (\acs{als}: \acl{als}, \acs{ms}: \acl{ms}, \acs{dmd}:
    \acl{dmd}, \acs{sma}: \acl{sma}, \acs{lis}: \acl{lis}).
    $+$: frequent, $-$: infrequent.}
    \label{tab:incidence}
\end{table}
\todo{include FA}
\todo{cite Abnormal Oculomotor Functions in Amyotrophic Lateral Sclerosis 10.3988/jcn.2018.14.4.464}
\todo{cite Neuro-Ophthalmologic Manifestations of Amyotrophic Lateral Sclerosis}

@@

To summarize: for patients with eye motor impairment, gazing directly at a screen target may
be uncomfortable, impractical, or even impossible.
Hence, assistive devices that rely on eye tracking are often inefficient for
them.
Consequently, while \ac{bci}s hold great promise for these individuals, conventional
gaze-dependent \ac{bci} solutions do not meet their needs due to the absence of gaze
control.
Therefore, the development of decoding strategies that account for covert
VSA becomes crucial in the pursuit of high-performance gaze-independent
\ac{bci}s.

In this work, we will use the term \emph{gaze-independent} meaning
\emph{‘dealing explicitly with the fact that a user cannot control their
gaze.'}
In the context of a visual \ac{bci}, this means that the user's
visuospatial attention and their gaze do not necessarily coincide.

@@

Gaze-independent \ac{erp}-based \ac{bci}s~\cite{Riccio2012, Aloise2012} can be realized in three
ways.
Firstly, active \ac{bci} communication paradigms relying on endogenous activation from the user
do not rely on sensory stimulation.
Examples of this are imagined movement or imagined speech paradigms.
These paradigms can yield very high information transfer
rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
complexity that can be captured in commands, but often only do so when paired
with invasive recording.
Non-visual reactive paradigms that use e.g. auditory and somatosensory
stimulation do not rely on gaze redirection exist but also result in lower information transfer
rates compared to visual paradigms.
Furthermore, they can suffer from increased mental effort in operation and
user-dependent variability~\cite{Severens2014,Reichert2020b}.
\todo{insert table from riccio in text as numbers, not as table}

@@

First, we designed a visual oddball interface that helps us study and adapt
to the effects of eye motor impairment on the \ac{erp}, when using existing state-of-the-art
decoders and on our own proposed decoders.
Using a hexagonal lay-out interface, similar to the visual Hex-o-Spell proposed
\textcite{Treder2010}, we present six flashing circular targets
(without letters or symbols) to the participant while the EEG and
electro-oculogram (EOG) are recorded, as well as eye gaze using eye tracking.
\todo{chosen because gaze-independent optimized and Peters2020}

### 2024-09-29 chapters/patients.tex
Add decoders
Intro, P300 stimulation
Started patients paper

\section{Patient presentation}
\section{Data collection \& preprocessing}
\todo{Define structure patient results/discussion}
\section{Outcomes}
\section{Discussion}

>>>>

\section{Introduction}

Communication BCIs have a target population consisting of patients in various
stages of paralysis or \ac{lis}, often suffering from oculomotor impairments.
This reduces their performance operating visual oddball \ac{bci}s, since they
are presumed unable to comfortably redirect their gaze at the desired target,
i.e., perform overt \ac{vsa}.
They might more naturally operate in covert \ac{vsa}, where the gaze and
\ac{vsa} do not coincide.
Several studies show that performance drops in covert VSA \todo{[2, 1], our
study}, necessitating gaze-independent solutions.
Our previous study, presented in chapter \todo{ref chapter}, indicates gaze-independent performance can be
improved in healthy subjects using a suited decoding strategy.

These studies build on the assumption that patients with eye motor impairment
would feel comfortable operating
Visual paradigms

Need for verification in patients, cite different characteristics, differences
in setting, environment and equipment,
\todo{goal is class 1, but challenging to perform research experiments and to
recruit, cite wolpaw}

problems eye tracking

Literature for visual \ac{bci} in eye motor impaired patients is sparse.
\todo{List conclusions of Peters2020 (severely eye motor impaired als patients
able to use a hex-o-spell)}

Therefore, we aim to apply the concepts from earlier work an literature to
patients with motor impairment and, in addition, eye motor impairment.
The objectives of this case study are as follows:
\begin{enumerate*}
  \item Explore capabilities and experienced comfort of eye motor impaired
    patients,
  \item evaluate the performance of a gaze-independent visual \ac{bci} in patients
  \item verify if this performance can be improved with a suitable decoding
    strategy.
\end{enumerate*}

\section{Materials \& methods}
\subsection{Patient recruitment}
Patients were recruited from across the Neuromuscular Reference center at
University Hospital Leuven (Leuven, Belgium), TRAINM Neuro Rehab Clinics
(Antwerp, Belgium), the Neurorehabilitation Unit at University Hospital Lille
(Lille, France) and a specialized care home (France).
Experiments were performed under the supervision of their treating physician.
Patients were recruited based on the following criteria.
To qualify for inclusion, patients must
\begin{enumerate}
	\item be at least 18 years old and no older than 60
	years,
  \item belong to class 2 or 3 according the \ac{bci}
	patient selection criteria presented by~\textcite{Wolpaw2006},\label{item:patients/inclusion/wolpaw}
  \todo{specific cutoff based on vision}
  \item have limitations to the extent or comfort of their eye motor control\label{item:patients/inclusion/oculomotor}
\end{enumerate}
Patients were excluded if they
\begin{enumerate}
  \item have a diagnosis of a major medical condition, including any major
    neurological or psychiatric disorder other than those of interest based on
    inclusion criteria~\ref{item:patients/inclusion/wolpaw}
    and~\ref{item:patients/inclusion/oculomotor}\label{item:patients/exclusion/medical}
  \item have a predisposition to or have a history of any kind of epileptic seizures,
    including photosensitive epilepsy,\label{item:patients/exclusion/epilepsy}
  \item have a severe loss in vision or hearing, that would significantly impair
        participation in the experiment,\label{item:patients/exclusion/vision}
  \item are currently using specific psychoactive medications or substances that could affect the out-
        come.\label{item:patients/exclusion/cognitive}
  \item be able to understand the experiment instructions and cooperate,
  \item have any other limitations preventing them from performing the given task.
\end{enumerate}

In total, 11 patients were contacted, of which 1 \ac{ms} patient was excluded based on
criterion~\ref{item:patients/exclusion/vision}, 1 \ac{tbi} patient on
both~\ref{item:patients/exclusion/epilepsy}
and~\ref{item:patients/exclusion/cognitive}, and one stroke patient based
on~\ref{item:patients/exclusion/medical}
One further stroke patient was excluded due to technical
difficulties during the experimental session.
Vision was assessed using a LogMAR chart~\cite{Bailey1976}.

Ultimately, 7 patients were retained.
Of these, one patient was diagnosed from bulbar-onset \ac{als}.
\ac{als} is a neurodegenerative disease affecting the motor neurons, leading to
progressive loss of motor function.
This initially results in general weakness and loss of muscle tone, but
eventually results in full body paralysis.\todo{cite}
Although speech and especially eye movements are usually preserved until the
later stages of the disease progression, the bulbar-onset variant is
characterized by an early loss of speech and an increased involvement of eye
motor symptoms~\cite{Guo2022}\todo{which?}.
Furthermore, one of the goals of \ac{bci} is to support \ac{als} patients whose
life span has been extended with life support with and assistive technology to
ensure quality of life.
In these very progressed stages of \ac{als}, eye movement will eventually also
be affected~\cite{Hayashi1991}.

Three other patients were diagnosed with \ac{fa}, a neurodegenerative
disease affecting the
spinal cord, peripheral nervous system and cerebellum.
\ac{fa} results in an impairing loss of muscle coordination.
In progressed stages, the disease can present with nystagmus, saccadic
intrusions and gaze dysmetria~\cite{Cook2017}.

The final three patients were stroke patients.
Brain stem and cerebellar stroke can often lead to forms of \ac{lis}, and
various eye motor disorders are common depending on the exact location
~\cite{Bogousslavsky1987, Moncayo2009}.
Stroke can lead to some of the more severe impairments like ophthalmoplegia or
partial ophthalmoplegia immediately from onset.

Table~\ref{tab:patients/patients} lists the included patients and their
diagnoses.

\begin{table}[t]
  \centering
  \footnotesize
  \input{chapters/patients_table.tex}
  \caption{Presentation of included patients including their diagnosis and
  capabilities. (Trach.: patient underwent a tracheotomy, Cls.: classification according to~\textcite{Wolpaw2006}).
  }
  \label{tab:patients/patients}
\end{table}

\subsection{Eye tracking and eye motor examination}

Self-reported eye motor abnormalities were recorded.
Patients were asked for, the paralysis or inability to move in a given
direction of one or both eyes, the inability to pursue an object to specific
positions, discomfort or fatigue fixating at specific positions or in general
when performing visual tasks and eye tremors while fixating, resting or
pursuing.

Additionaly, we implemented and performed the automated NeuroEye eye movement
test proposed by~\textcite{Hassan2022}.
This test is based on calibration-free eye tracking.
\todo{say which part we used and which part not}
\todo{motivation and text and references Juliette}
\todo{Limited in what it can tell us (check paper)}
\todo{include eye test document juliette}
Table~\ref{tab:patients/eye} details their eye motor impairments and vision of
included patients.
Patient PA1 had the mildest impairment, only reporting fatigue when fixating
for prolonged times.
Eye motor function of subjects PC2 and PC3 were most severly affected.
Patient PC2 was only able to look up and down, and patient PC3 only retained
partial movement of the right eye, while the left eye was permanently closed.

\begin{table}[t]
  \centering
  \footnotesize
  \input{chapters/patients_table_eye.tex}
  \caption{Vision and eye motor impairment for included patients.
  Eye motor impairment was assessed with a combination of self-reported issues
  by the subject and the NeuroEye~\cite{Hasan2022} test.
  LogMAR: lower is better.}
  \label{tab:patients/eye}
\end{table}

Eye tracking throughout the experimental session was performed using the Tobii~\todo{model, make}
portable eye tracker.
\todo{describe analysis during experiment}

\subsection{\Ac{bci} stimulation}

The \ac{bci} stimulation procedure was based on the
Hex-o-Spell~\cite{Treder2010} implementation presented
by~\textcite{VanDenKerchove2024}.
Similar to this study, the task consists of counting the flashes of a cued
target among all flashing targets.
We refer to section~\ref{sec:covert-align/stimulation} for the exact details.

To adapt the interface for the patient setting, the number of blocks was
decreased to 6 per \ac{vsa} setting.
An additional \ac{free \ac{vsa}} condition was introduced.
Here, the patient was instructed to perform the task as they deemed most
comfortable.
The cued \emph{split attention} setting proposed
by~\textcite{VanDenKerchove2024} is not studied here, since we are interested
in natural \ac{vsa} operation settings for gaze-impaired patients.
If the patient was not fully paralyzed, they were instructed not to move the head.
\Ac{isi} was increased to\tod{number} to decrease task difficulty.
The experiment alsy started with a training block in each condition, where the
patient was instructed with feedback on their performance to make sure they
understood and were able to perform the task.

\fourpanefig{%
  \figpane{\includegraphics[width=\textwidth]{figures/patients/PD01a-obfuscated.jpg}}{A
  patient with the stimulation and recording setup.}{fig:patients/photos/side}
}{%
  \figpane{\includegraphics[width=\textwidth]{figures/patients/PD01b-obfuscated.jpg}}{subcaption}{fig:patients/photos/front}
}{%
  the interface
}{%
  4
}{%
  caption
}{fig:patients/photos}

Mention change in ISI
Mention less rounds

\subsection{Data collection \& preprocessing}

\todo{check checklist}
\subsection{BCI Decoding}

\section{Outcomes}

\subsection{Eye tracking analysis}

PC2 reported that he was unable to fixate the lower two targets, and th

\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/patients/fig_gaze.png}
}{%
  Distribution of the recorded gaze position during the experimental session in the three \ac{vsa}
  conditions.
  Subjects PB2 and PC4 preferred covert \ac{bci} operation, with PB2 resting gaze
  near the middle of the screen, and PC4 near the bottom.
}{%
  fig:patients/gaze
}

\subsection{Decoding performance}

\begin{figure}[t]
  \input{figures/patients/fig_decode.pgf}
\end{figure}
\todo{check if percentages or 0-1 accuracy and roc-auc}

For PB2 and PC4 performing covertly, was free decoding higher than covert
decoding?

\subsection{Cross-condition calibration}

| indication of fatigue, cinsistency within free case
| closer to overt or to covert by looking at eye tracker
|
| will learn
| split up per target
|
| Findings:
| 1. Some patients prefer to do covert (check performance results for them),
|    analyze distance to target
| 2. covert attention can be improved using WCBLE for all subjects, but not to
| the
|    level of overt covert attention works best when training in overt
| attention
| s this also true for patients preferring covert?
| 3. free performs on par (in all subjects?), opens debate of necessity of
|    gaze-independence BCIs
| 4. covert attention works best when training in overt attention. Is this also
|    true for patients preferring covert?

2024-09-18 04:39:35 PM can more overt attention epochs closer to
| mention problem can't know if patient shifted or eye tracker shifted
| (difficult mobile gaze tracker)
| overt lower because can't perform

\section{Discussion}

patients 2 are progressing towards 3

Limitations

Limited sample of patients, FA maybe not most interesting group to represent
real BCI users

interpreting eye tracker results hard because of inaccuracy eye tracker for eye
motor impaired patients. Redo with more advanced stationary eye tracker and
head rest.

advanced eye tracking equipment is necessary

user comfort was not objectively measured using established questionnaires

Further work
Comparing with a sample of non-gaze impaired patients with  neuromuscular
disability DMD
On-line session with these patients

In a follow up study, we aim to revisit these patients with an on-line system
to be able to calculate a realistic \ac{itr} and measure their user experience.

\section{Conclusion}

\todo{https://sci-hub.ru/https://ieeexplore.ieee.org/abstract/document/1642754}
\todo{https://www.sciencedirect.com/science/article/abs/pii/B9780444639349000020}
\todo{BCI principles and practice chapter 19}
\todo{https://www.nature.com/articles/s44222-024-00239-5}
